\pstart
\eledsection*{I}

\pend\pstart
In the records of the twelve tribes of Israel was Joachim, a man rich exceedingly; and he brought his offerings double\footnote{Susanna i. 4.}, saying: There shall be of my superabundance to all the people, and there shall be the offering for my forgiveness\footnote{The readings vary, and the sense is doubtful. Thilo thinks that the sense is: What I offer over and above what the law requires is for the benefit of the whole people; but the offering I make for my own forgiveness (according to the law's requirements) shall be to the Lord, that He may be rendered merciful to me.} to the Lord for a propitiation\footnote{The Church of Rome appoints March 20 as the Feast of St. Joachim. His liberality is commemorated in prayers, and the lessons to be read are Wisd. xxxi. and Matt. i.} for me.

\pend\pstart
For the great day of the Lord was at hand, and the sons of Israel were bringing their offerings. And there stood over against him Rubim, saying: It is not meet for thee first to bring thine offerings, because thou hast not made seed in Israel\footnote{1 Sam. i. 6, 7; Hos. ix. 14.}.

\pend\pstart
And he searched, and found that all the righteous had raised up seed in Israel. And he called to mind the patriarch Abraham, that in the last day\footnote{Another reading is: In his last days.} God gave him a son Isaac.

\pend\pstart
And Joachim was exceedingly grieved, and did not come into the presence of his wife; but he retired to the desert\footnote{Another reading is: Into the hill-country.}, and there pitched his tent, and fasted forty days and forty nights,\footnote{Moses: Ex. xxiv. 18, xxxiv. 28; Deut. ix. 9. Elijah: 1 Kings xix. 8. Christ: Matt. iv. 2.} saying in himself: I will not go down either for food or for drink until the Lord my God shall look upon me, and prayer shall be my food and drink.

\pend\pstart
\eledsection*{II}

\pend\pstart
And his wife Anna\footnote{The 26th day of July is the Feast of St. Anna in the Church of Rome.} mourned in two mournings, and lamented in two lamentations, saying: I shall bewail my widowhood; I shall bewail my childlessness.

\pend\pstart
And the great day of the Lord was at hand; and Judith\footnote{Other forms of the name are Juth, Juthin.} her maid-servant said: How long dost thou humiliate thy soul? Behold, the great day of the Lord is at hand, and it is unlawful for thee to mourn. But take this head-band, which the woman that made it gave to me; for it is not proper that I should wear it, because I am a maid-servant, and it has a royal appearance\footnote{Some MSS. have: For I am thy maid-servant, and thou hast a regal appearance.}.

\pend\pstart
And Anna said: Depart from me; for I have not done such things, and the Lord has brought me very low. I fear that some wicked person has given it to thee, and thou hast come to make me a sharer in thy sin. And Judith said: Why should I curse thee, seeing that\footnote{Several MSS. insert: Thou hast not listened to my voice; for.} the Lord hath shut thy womb, so as not to give thee fruit in Israel?

\pend\pstart
And Anna was grieved exceedingly, and put off her garments of mourning, and cleaned her head, and put on her wedding garments, and about the ninth hour went down to the garden to walk. And she saw a laurel, and sat under it, and prayed to the Lord, saying: O God of our fathers, bless me and hear my prayer, as Thou didst bless the womb of Sarah, and didst give her a son Isaac\footnote{Comp. 1 Sam. i. 9\textendash 18.}.

\pend\pstart
\eledsection*{III}

\pend\pstart
And gazing towards the heaven, she saw a sparrow's nest in the laurel\footnote{Tobit ii. 10.}, and made a lamentation in herself, saying: Alas! who begot me? and what womb produced me? because I have become a curse in the presence of the sons of Israel, and I have been reproached, and they have driven me in derision out of the temple of the Lord.

\pend\pstart
Alas! to what have I been likened? I am not like the fowls of the heaven, because even the fowls of the heaven are productive before Thee, O Lord. Alas! to what have I been likened? I am not like the beasts of the earth, because even the beasts of the earth are productive before Thee, O Lord.

\pend\pstart
Alas! to what have I been likened? I am not like these waters, because even these waters are productive before Thee, O Lord. Alas! to what have I been likened? I am not like this earth, because even the earth bringeth forth its fruits in season, and blesseth Thee, O Lord\footnote{Many of the MSS. here add: Alas! to what have I been likened? I am not like the waves of the sea, because even the waves of the sea, in calm and storm, and the fishes in them, bless Thee, O Lord.}.

\pend\pstart
\eledsection*{IV}

\pend\pstart
And, behold, an angel of the Lord stood by, saying: Anna, Anna, the Lord hath heard thy prayer, and thou shalt conceive, and shall bring forth; and thy seed shall be spoken of in all the world. And Anna said: As the Lord my God liveth, if I beget either male or female, I will bring it as a gift to the Lord my God; and it shall minister to Him in holy things all the days of its life\footnote{1 Sam. i. 11.}.

\pend\pstart
And, behold, two angels came, saying to her: Behold, Joachim thy husband is coming with his
flocks\footnote{One of the MSS.: With his shepherds, and sheep, and goats, and oxen.}. For an angel of the Lord went down to him, saying: Joachim, Joachim, the Lord God hath heard thy prayer. Go down hence; for, behold, thy wife Anna shall conceive.

\pend\pstart
And Joachim went down and called his shepherds, saying: Bring me hither ten she-lambs without spot or blemish, and they shall be for the Lord my God; and bring me twelve tender calves, and they shall be for the priests and the elders; and a hundred goats for all the people.

\pend\pstart
And, behold, Joachim came with his flocks; and Anna stood by the gate, and saw Joachim coming, and she ran and hung upon his neck, saying: Now I know that the Lord God hath blessed me exceedingly; for, behold the widow no longer a widow, and I the childless shall conceive. And Joachim rested the first day in his house.

\pend\pstart
\eledsection*{V}

\pend\pstart
And on the following day he brought his offerings, saying in himself: If the Lord God has been rendered gracious to me, the plate\footnote{Ex. xxviii. 36\textendash 38. For traditions about the \textit{petalon}, see Euseb., \textit{H. E.}, ii. 23, iii. 31, v. 24; Epiph., \textit{Hær.}, 78.} on the priest's forehead will make it manifest to me. And Joachim brought his offerings, and observed attentively the priest's plate when he went up to the altar of the Lord, and he saw no sin in himself. And Joachim said: Now I know that the Lord has been gracious unto me, and has remitted all my sins. And he went down from the temple of the Lord justified, and departed to his own house.

\pend\pstart
And her months were fulfilled, and in the ninth\footnote{Various readings are: Sixth, seventh, eighth.} month Anna brought forth. And she said to the midwife: What have I brought forth? and she said: A girl. And said Anna: My soul has been magnified this day. And she laid her down. And the days having been fulfilled, Anna was purified, and gave the breast to the child\footnote{One of the MSS. inserts: On the eighth day.}, and called her name Mary.

\pend\pstart
\eledsection*{VI}

\pend\pstart
And the child grew strong day by day; and when she was six\footnote{One of the MSS. has nine.} months old, her mother set her on the ground to try whether she could stand, and she walked seven steps and came into her bosom; and she snatched her up, saying: As the Lord my God liveth, thou shalt not walk on this earth until I bring thee into the temple of the Lord. And she made a sanctuary in her bed-chamber, and allowed nothing common or unclean to pass through her. And she called the undefiled daughters of the Hebrews, and they led her astray\footnote{This is the reading of most MSS.; but it is difficult to see any sense in it. One MS. reads: They attended on her. Fabricius proposed: They bathed her.}.

\pend\pstart
And when she was a year old, Joachim made a great feast, and invited the priests, and the scribes, and the elders, and all the people of Israel. And Joachim brought the child to the priests; and they blessed her, saying: O God of our fathers, bless this child, and give her an everlasting name to be named in all generations. And all the people said: So be it, so be it, amen. And he brought her to the chief priests; and they blessed her, saying: O God most high, look upon this child, and bless her with the utmost blessing, which shall be for ever.

\pend\pstart
And her mother snatched her up, and took her into the sanctuary of her bed-chamber, and gave her the breast. And Anna made a song to the Lord God, saying: I will sing a song to the Lord my God, for He hath looked upon me, and hath taken away the reproach of mine enemies; and the Lord hath given the fruit of His righteousness, singular in its kind, and richly endowed before Him. Who will tell the sons of Rubim that Anna gives suck? Hear, hear, ye twelve tribes of Israel, that Anna gives suck. And she laid her to rest in the bed-chamber of her sanctuary, and went out and ministered unto them. And when the supper was ended, they went down rejoicing, and glorifying the God of Israel\footnote{Two of the MSS. add: And they gave her the name of Mary, because her name shall not fade forever. This derivation of the name\textemdash from the root \textit{mar}, fade\textemdash is one of a dozen or so.}.

\pend\pstart
\eledsection*{VII}

\pend\pstart
And her months were added to the child. And the child was two years old, and Joachim said: Let us take her up to the temple of the Lord, that we may pay the vow that we have vowed, lest perchance the Lord send to us\footnote{This is taken to mean: Send someone to us to warn us that we have been too long in paying our vow. One MS. reads, lest the Lord depart from us; another, lest the Lord move away from us.}, and our offering be not received. And Anna said: Let us wait for the third year, in order that the child may not seek for father or mother. And Joachim said: So let us wait.

\pend\pstart
And the child was three years old, and Joachim said: Invite the daughters of the Hebrews that are undefiled, and let them take each a lamp, and let them stand with the lamps burning, that the child may not turn back, and her heart be captivated from the temple of the Lord. And they did so until they went up into the temple of the Lord. And the priest received her, and kissed her, and blessed her, saying: The Lord has magnified thy name in all generations. In thee, on the last of the days, the Lord will manifest His redemption to the sons of Israel.

\pend\pstart
And he set her down upon the third step of the altar, and the Lord God sent grace upon her; and she danced with her feet, and all the house of Israel loved her.

\pend\pstart
\eledsection*{VIII}

\pend\pstart
And her parents went down marvelling, and praising the Lord God, because the child had not turned back. And Mary was in the temple of the Lord as if she were a dove that dwelt there, and she received food from the hand of an angel.

\pend\pstart
And when she was twelve\footnote{Or, fourteen. Postel's Latin version has \textit{ten}.} years old there was held a council of the priests, saying: Behold, Mary has reached the age of twelve years in the temple of the Lord. What then shall we do with her, lest perchance she defile the sanctuary of the Lord? And they said to the high priest: Thou standest by the altar of the Lord; go in, and pray concerning her; and whatever the Lord shall manifest unto thee, that also will we do.

\pend\pstart
And the high priest went in, taking the robe\footnote{Ex. xxviii. 28; Sirach xlv. 9; Justin, \textit{Tryph}., xlii.} with the twelve bells into the holy of holies; and he prayed concerning her. And behold an angel of the Lord stood by him, saying unto him: Zacharias, Zacharias, go out and assemble the widowers of the people, and let them bring each his rod; and to whomsoever the Lord shall show a sign, his wife shall she be. And the heralds went out through all the circuit of Judæa, and the trumpet of the Lord sounded, and all ran.

\pend\pstart
\eledsection*{IX}

\pend\pstart
And Joseph, throwing away his axe, went out to meet them; and when they had assembled, they went away to the high priest, taking with them their rods. And he, taking the rods of all of them, entered into the temple, and prayed; and having ended his prayer, he took the rods and came out, and gave them to them: but there was no sign in them, and Joseph took his rod last; and, behold, a dove came out of the rod, and flew upon Joseph's head. And the priest said to Joseph, Thou hast been chosen by lot to take into thy keeping the virgin of the Lord.

\pend\pstart
But Joseph refused, saying: I have children, and I am an old man, and she is a young girl. I am afraid lest I become a laughing-stock to the sons of Israel. And the priest said to Joseph: Fear the Lord thy God, and remember what the Lord did to Dathan, and Abiram, and Korah\footnote{Num. xvi. 31\textendash 33.}; how the earth opened, and they were swallowed up on account of their contradiction. And now fear, O Joseph, lest the same things happen in thy house.

\pend\pstart
And Joseph was afraid, and took her into his keeping. And Joseph said to Mary: Behold, I have received thee from the temple of the Lord; and now I leave thee in my house, and go away to build my buildings, and I shall come to thee. The Lord will protect thee.

\pend\pstart
\eledsection*{X}

\pend\pstart
And there was a council of the priests, saying: Let us make a veil for the temple of the Lord. And the priest said: Call to me the undefiled virgins of the family of David. And the officers went away, and sought, and found seven virgins. And the priest remembered the child Mary, that she was of the family of David, and undefiled before God. And the officers went away and brought her. And they brought them into the temple of the Lord. And the priest said: Choose for me by lot who shall spin the gold, and the white\footnote{Lit., undefiled. It is difficult to say what colour is meant, or if it is a colour at all. The word is once used to mean the sea, but with no reference to colour. It is also the name of a stone of a greenish hue.}, and the fine linen, and the silk, and the blue\footnote{Lit., hyacinth.}, and the scarlet, and the true purple\footnote{Ex. xxv. 4.}. And the true purple and the scarlet fell to the lot of Mary, and she took them, and went away to her house. And at that time Zacharias was dumb, and Samuel was in his place until the time that Zacharias spake. And Mary took the scarlet, and span it.

\pend\pstart
\eledsection*{XI}

\pend\pstart
And she took the pitcher, and went out to fill it with water. And, behold, a voice saying: Hail, thou who hast received grace; the Lord is with thee; blessed art thou among women!\footnote{Luke i. 28.} And she looked round, on the right hand and on the left, to see whence this voice came. And she went away, trembling, to her house, and put down the pitcher; and taking the purple, she sat down on her seat, and drew it out.

\pend\pstart
And, behold, an angel of the Lord stood before her, saying: Fear not, Mary; for thou hast found grace before the Lord of all, and thou shalt conceive, according to His word. And she hearing, reasoned with herself, saying: Shall I conceive by the Lord, the living God? and shall I bring forth as every woman brings forth?

\pend\pstart
And the angel of the Lord said: Not so, Mary; for the power of the Lord shall overshadow thee: wherefore also that holy thing which shall be born of thee shall be called the Son of the Most High. And thou shalt call His name Jesus, for He shall save His people from their sins. And Mary said: Behold, the servant of the Lord before His face: let it be unto me according to thy word.

\pend\pstart
\eledsection*{XII}

\pend\pstart
And she made the purple and the scarlet, and took them to the priest. And the priest blessed her, and said: Mary, the Lord God hath magnified thy name, and thou shalt be blessed in all the generations of the earth.

\pend\pstart
And Mary, with great
joy, went away to Elizabeth her kinswoman,\footnote{Luke i. 39, 40.} and knocked at the door. And when Elizabeth heard her, she threw away the scarlet,\footnote{Other readings are: the wool\textemdash what she had in her hand.} and ran to the door, and opened it; and seeing Mary, she blessed her, and said: Whence is this to me, that the mother of my Lord should come to me? for, behold, that which is in me leaped and blessed thee.\footnote{Luke i. 34, 44.} But Mary had forgotten the mysteries of which the archangel Gabriel had spoken, and gazed up into heaven, and said: Who am I, O Lord, that all the generations of the earth should bless me?\footnote{Luke i. 48.}

\pend\pstart
And she remained three months with Elizabeth; and day by day she grew bigger. And Mary being afraid, went away to her own house, and hid herself from the sons of Israel. And she was sixteen\footnote{Six MSS. have \textit{sixteen}; one, \textit{fourteen}; two, \textit{fifteen}; and one, \textit{seventeen}.} years old when these mysteries happened.

\pend\pstart
\eledsection*{XIII}

\pend
